---
title: "Risor ruber - Genetic analyses: relatedness and basic popgen"
date: "2026-08-12"
---

This code refers to Bunholi et al. 2026 "Ecological constraints of habitat specialization shape the social organization and life-history of a sponge-dwelling goby."

## Analysis overview

This report reproduces the population-genomic, spatial genetic-distance, and relatedness analyses for the ddRADseq dataset. The population-structure dataset contains 42 biological samples and two technical replicates; the relatedness dataset contains 33 retained individuals.
Manuscript figures: 2A-D
Supplementary figures: S6 and S7

### Required inputs

All inputs are provided in `data/relatedness_popgen/`:

- "pop_m5_r07_wl_biallelic_mac5.noHWEexcess_p1e4_LDpruned.vcf" (population structure vcf)

- "rel_m5_r095_wl_biallelic_mac16.noHWEexcess_p1e4.LDpruned.vcf" (relatedness vcf)

- "Risor_ruber_2023_Pairwise_Distances.csv" (pairwise geographic distances)

- "Risor_rel_pan_1664SNPs_withDist.csv" (dyad distances + relatedness data)

- "Risor_metadata_geno_subset_with_all_variables.csv" (metadata all)

User: please run this R Markdown file from the Scripts/ directory.

### Software setup

The following chunk lists the packages required by this report. 

```{r setup-packages}
# Install required packages before the first run.
# Uncomment these commands only if the packages are not already installed.

# install.packages(c(
#   "tidyverse", "ggrepel", "lme4", "lmerTest", "performance",
#   "vcfR", "hierfstat", "adegenet", "BiocManager", "remotes"
# ))

# SNPRelate is installed through Bioconductor:
# BiocManager::install("SNPRelate")

# radiator is installed from GitHub:
# remotes::install_github("thierrygosselin/radiator")

#load packages
pkgs <- c(
  "tidyverse", "ggrepel", "lme4", "lmerTest", "SNPRelate",
  "performance", "vcfR", "hierfstat", "adegenet", "radiator"
)
invisible(lapply(pkgs, require, character.only = TRUE))
```

### Data preparation

This section defines project relative paths, imports the two VCF datasets and associated tabular data, and standardizes identifiers used for individual and dyad matching. Generated intermediate files are written to `outputs/`.

```{r}
# Run this R Markdown file from the Scripts/ directory. All required inputs
# are read from data/relatedness_popgen/ and generated files go to outputs/.
data_dir <- file.path("..", "data", "relatedness_popgen")
output_dir <- file.path("..", "outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

vcf_all <- file.path(data_dir, "pop_m5_r07_wl_biallelic_mac5.noHWEexcess_p1e4_LDpruned.vcf")
vcf_rel <- file.path(data_dir, "rel_m5_r095_wl_biallelic_mac16.noHWEexcess_p1e4.LDpruned.vcf")
dist <- read.csv(file.path(data_dir, "Risor_ruber_2023_Pairwise_Distances.csv"))
rel_df <- read.csv(file.path(data_dir, "Risor_rel_pan_1664SNPs_withDist.csv"))
gds_out <- file.path(output_dir, "pop_m5_r07_wl_biallelic_mac5.noHWEexcess_p1e4_LDpruned.vcf.gds")

meta_all <- read_csv(file.path(data_dir, "Risor_metadata_geno_subset_with_all_variables.csv")) #overall metadata
meta_all <- meta_all %>%
  mutate(sponge_id = if_else(sponge_id == "GOL_06", "GOL_6", sponge_id))
sponge_meta <- meta_all %>%
  dplyr::select(id, sponge_id, S_width_cm, Reef, S_species)
```

```{r}
#helper functions
std_id <- function(x) x |> stringr::str_trim() |> gsub("-", "_", x = _) #standardizing names
std_pair <- function(a, b) if_else(a < b, paste(a,b,sep="__"), paste(b,a,sep="__")) #standardizing dyad direction
```

### Prepare the COANCESTRY input

The relatedness VCF is converted to the allele-coded format required by COANCESTRY. The expected output is `outputs/new_coancestry_input_filtered_dataset_related.txt`; relatedness estimation itself is performed in COANCESTRY outside R (Windows GUI software https://www.zsl.org/about-zsl/resources/software/coancestry)

```{r}
related_data<- tidy_vcf(vcf_rel) #filtered dataset

#Check what genotype columns exist
names(related_data)
```

```{r}
#VCF generated from STACKS does not have a GT column, but GT_BIN
#GT_BIN: 0 = homozygous reference (0/0); 1 = heterozygous (0/1); 2 = homozygous alternative (1/1); NA = missing data

#Create GT column based on GT_BIN
related_data$GT <- dplyr::case_when(
  related_data$GT_BIN == 0 ~ "001001",  #homozygous reference
  related_data$GT_BIN == 1 ~ "001002",  #heterozygous
  related_data$GT_BIN == 2 ~ "002002",  #homozygous alternate
  TRUE ~ "000000"                       #missing data
)

#writing Coancestry input
radiator::write_related(
  data = related_data,
  filename = file.path(output_dir, "new_coancestry_input_filtered_dataset.txt"))
```

#### Compare relatedness estimators

The deposited COANCESTRY results contain several estimators. Their Spearman correlations are compared here to document estimator agreement and support use of the Wang estimator in the downstream models.
rel_df is the output from COANCESTRY. The code above is only to instruct how to generate COANCESTRY input file from a VCF since it can be really tricky; however, it is not needed to reproduce the results since the relatedness estimations are already being provided.

```{r}
rel_all <- rel_df %>%
  transmute(
    Individual1 = std_id(Individual1),
    Individual2 = std_id(Individual2),
    pair = std_pair(std_id(Individual1), std_id(Individual2)),
    Distance_m,
    Wang, Ritland, LynchRd, LynchLi, QuelletGt, TrioML
  )
```

```{r}
estimators <- c("Wang","Ritland","LynchRd","QuelletGt")

cor_mat <- cor(rel_all[,estimators], method="spearman", use="pairwise.complete.obs") #generating correlation matrix

as.data.frame(as.table(cor_mat)) %>%
  ggplot(aes(Var1, Var2, fill=Freq)) +
  geom_tile() +
  theme_minimal() +
  labs(x=NULL, y=NULL, fill="Spearman ρ") +
  coord_equal()
```

All p values are high (~0.8–1.0), showing strong correlation between metrics. Wang was then chosen because it performs well with moderate SNP numbers (Wang, 2002).

## 1. Relatedness analyses

In this section, the pairwise Wang relatedness is modeled as a function of geographic distance and environmental dissimilarity. 
The expected outputs are model summaries and manuscript Figure 2A-B.

```{r}
hist(rel_df$Wang)
```

```{r}
#arranging relatdness df and adding other metrics from metadata
rel_df2 <- rel_df %>%
  transmute(
    Individual1 = std_id(Individual1),
    Individual2 = std_id(Individual2),
    Wang,
    Distance_m) %>%
  left_join(
    meta_all %>%
      transmute(
        Individual1 = std_id(id),
        sp1 = sponge_id,
        size1 = S_width_cm,
        reef1 = Reef,
        species1 = S_species,
        is_male1 = is_male),
    by = "Individual1") %>%
  left_join(
    meta_all %>%
      transmute(
        Individual2 = std_id(id),
        sp2 = sponge_id,
        size2 = S_width_cm,
        reef2 = Reef,
        species2 = S_species,
        is_male2 = is_male
      ),
    by = "Individual2"
  ) %>%
  mutate(
    same_sponge    = sp1 == sp2,
    same_species   = species1 == species2,
    diff_sponge_sz = abs(size1 - size2),
    same_reef      = reef1 == reef2,
    same_sex       = is_male1 == is_male2,
    distance_km    = Distance_m / 1000,
    log_dist       = log1p(distance_km),
    i              = factor(Individual1),
    j              = factor(Individual2)
  )

```

```{r}
##LMM: Wang ~ geo_dist + env metrics (relatedness dataset: 33 indv = 528 dyads)
m_rel <- lmer(Wang ~ log_dist + same_sponge + diff_sponge_sz + same_reef + same_species + same_sex +
                (1|i) + (1|j),
             data = rel_df2, REML = TRUE)
summary(m_rel)
```

```{r}
#Checking model fit 
model_performance(m_rel)
check_model(m_rel)
check_collinearity(m_rel)
isSingular(m_rel) #small but still present
r2(m_rel) #fixed effects explain less than 1% of genetic distance
```

Interpretation:

- Relatedness is centered and scaled relative to allele frequencies, so it's more sensitive to subtle kinship patterns and therefore lots of variance is attributed to individual identity (dyad structure) whereas only 4.5% is explained by fixed effects
- Some individuals are generally more related to others than average (background structure), but not in a way predicted by sponge/reef variables.
- Relatedness patterns are not ecologically structured at the microhabitat scale.

### Visualize relatedness

Kin structure - are individuals sharing a sponge more related than everyone else?

```{r}
#arranging df
rel_df2 <- rel_df2 %>%
  mutate(
    dyad_loc = case_when(
      same_sponge ~ "same group",          #same sponge
      same_reef   ~ "same reef",           #different sponge, same reef
      TRUE        ~ "different reefs"
    ),
    dyad_loc = factor(
      dyad_loc,
      levels = c("different reefs", "same reef", "same group")
    )
  )

means_df_rel <- rel_df2 %>%
  dplyr::group_by(dyad_loc) %>%
  dplyr::summarise(
    mean_wang = mean(Wang, na.rm = TRUE),
    .groups = "drop"
  )
```

```{r}
#Figure 2A - Relatedness distribution

cols_rel <- c(
  "same group"       = "grey70",
  "same reef"       = "skyblue2",
  "different reefs" = "skyblue4")

ggplot(rel_df2, aes(x = Wang, fill = dyad_loc)) +
  geom_density(alpha = 0.7, adjust = 1.0, color = NA) +
  geom_vline(
    data = means_df_rel,
    aes(xintercept = mean_wang, colour = dyad_loc),
    linewidth = 0.7,
    linetype = "dashed",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = cols_rel,
    name   = "Dyad location"
  ) +
  scale_colour_manual(
    values = cols_rel
  ) +
  labs(
    x = "Relatedness (Wang estimator)",
    y = "Density"
  ) +
  theme_classic() +
  theme(
    legend.position      = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background    = element_rect(fill = "white", colour = NA)
  )

ggsave(file.path(output_dir, "Fig_2A.pdf"), width = 5, height = 4, units = "in")
```

```{r}
#Figure 2B - Relatedness by distance class

#arranging df
rel_df2 <- rel_df2 %>%
  mutate(
    distance_km = if_else(!is.na(distance_km),
                          distance_km,
                          Distance_m / 1000),
    dist_bin = cut(
      distance_km,
      breaks = c(0, 0.025, 0.075, 0.2, 3.5, Inf),  # adjust if needed
      labels = c("≤25 m", "25–75 m", "75–200 m",
                 "200–3500 m", ">3500 m"),
      include.lowest = TRUE
    )
  )

bin_summ_rel <- rel_df2 %>%
  group_by(dist_bin) %>%
  summarise(
    n_dyads = n(),
    mean_wang = mean(Wang, na.rm = TRUE),
    se_wang   = sd(Wang, na.rm = TRUE) / sqrt(n()),
    .groups   = "drop"
  )

#plotting
ggplot(rel_df2, aes(x = dist_bin, y = Wang)) +
  # raw points
  geom_jitter(
    width  = 0.15,
    size   = 2,
    alpha  = 0.35,
    color  = "dodgerblue4"
  ) +
  #bin means
  geom_point(
    data        = bin_summ_rel,
    aes(x = dist_bin, y = mean_wang),
    inherit.aes = FALSE,
    size        = 3,
    color       = "black",
    fill        = "black",
    shape       = 21,
    stroke      = 0.8
  ) +
  #error bars
  geom_errorbar(
    data        = bin_summ_rel,
    aes(x = dist_bin,
        ymin = mean_wang - se_wang,
        ymax = mean_wang + se_wang),
    inherit.aes = FALSE,
    width       = 0.15,
    color       = "black"
  ) +
  labs(
    x = "Dyad distance category",
    y = "Relatedness (Wang)"
  ) +
  theme_classic() +
  theme(
    axis.text.x      = element_text(size = 10),
    axis.text.y      = element_text(size = 10),
    strip.background = element_blank(),
    legend.position  = "none"
  )

ggsave(file.path(output_dir, "Fig_2B.pdf"), width = 5, height = 4, units = "in")
```

## 2. Pairwise genetic-distance analyses

Identity-by-state (IBS) is calculated from the population VCF and converted to genetic distance (`1 - IBS`). 
The GDS file is generated from the deposited VCF and is treated as a reproducible intermediate.

```{r}
# Convert the deposited VCF to GDS when needed, then open it for SNPRelate.The GDS file is a generated intermediate file 
if (!file.exists(gds_out)) {
  snpgdsVCF2GDS(
    vcf.fn = vcf_all,
    out.fn = gds_out,
    method = "copy.num.of.ref"
  )
}
snpgdsSummary(gds_out)
genofile_all <- snpgdsOpen(gds_out)

#Computing IBS (identical-by-state) to extract genetic distance
ibs    <- snpgdsIBS(genofile_all, autosome.only = FALSE)
ibsnum <- snpgdsIBSNum(genofile_all, autosome.only = FALSE) #counts
samples <- ibs$sample.id

#Obtaining genetic distance from IBS
IBS<- ibs$ibs
diag(IBS) <- 1

Nmat<- ibsnum$ibs0 + ibsnum$ibs1 + ibsnum$ibs2 #number of loci shared instead of proportion

D <- 1 - IBS #calculating genetic distance matrix 
rownames(D) <- colnames(D) <- samples
rownames(IBS) <- colnames(IBS) <- samples
rownames(Nmat)   <- colnames(Nmat)   <- samples

```

### Assemble and model pairwise genetic distance

Genetic-distance dyads are matched to geographic and sponge metadata. Parallel models are fitted for the complete sample set and the filtered main-lineage set; 
their summaries and diagnostics provide the statistical results underlying Figure 2C-D.

```{r}
#Preparing data - cleaning IBS matrix
id_map <- meta_all %>%
  mutate(
    ind_id = str_match(id, ".*_(\\d+)(?:_d)?$")[,2] |> as.integer(),  # 52 from golden_52 or golden_52_d
    is_rep = str_detect(id, "_d$")
  ) %>%
  arrange(ind_id, is_rep) %>%                 #prefer non-_d if duplicated
  group_by(ind_id, sponge_id) %>% slice(1) %>% ungroup() %>%
  transmute(ind_id, sponge_id, sample = id)

keep_samples <- id_map$sample

IBS_clean <- IBS[keep_samples, keep_samples]

#Tidy IBS to one row per dyad
pairs_ibs <- IBS_clean %>%
  as.data.frame() %>%
  rownames_to_column("Individual1") %>%
  pivot_longer(-Individual1, names_to = "Individual2", values_to = "IBS")

#Tidy n_loci in the same way and join
pairs_n <- Nmat %>%
  as.data.frame() %>%
  rownames_to_column("Individual1") %>%
  pivot_longer(-Individual1, names_to = "Individual2", values_to = "n_loci")

pairs_ibs <- pairs_ibs %>%
  left_join(pairs_n, by = c("Individual1","Individual2")) %>%
  # keep one triangle only
  filter(Individual1 < Individual2) %>%
  mutate(
    gen_dist   = 1 - IBS,
    canon_pair = paste(pmin(Individual1, Individual2),
                       pmax(Individual1, Individual2), sep = "__")
  )

head(pairs_ibs)
```

```{r}
#Preparing data - pairwise geo distance 
dist2 <- dist %>%
  dplyr::mutate(
    a_id = pmin(IN_ind_id, NEAR_ind_id),
    b_id = pmax(IN_ind_id, NEAR_ind_id),
    a_sp = if_else(IN_ind_id <= NEAR_ind_id, IN_sponge_id, NEAR_sponge_id),
    b_sp = if_else(IN_ind_id <= NEAR_ind_id, NEAR_sponge_id, IN_sponge_id)
  ) %>%
  distinct(a_id, b_id, a_sp, b_sp, .keep_all = TRUE) %>%
  left_join(id_map %>% rename(a_id = ind_id, a_sp = sponge_id, s1 = sample),
            by = c("a_id","a_sp")) %>%
  left_join(id_map %>% rename(b_id = ind_id, b_sp = sponge_id, s2 = sample),
            by = c("b_id","b_sp")) %>%
  dplyr::mutate(
    Individual1 = pmin(s1, s2),
    Individual2 = pmax(s1, s2),
    canon_pair  = paste(Individual1, Individual2, sep = "__")
  ) %>%
  dplyr::select(canon_pair, Individual1, Individual2, Distance_m) %>%
  distinct(canon_pair, .keep_all = TRUE)

dist2 <- dist2 %>%
  filter(!is.na(Individual1), !is.na(Individual2)) #filtering to only pairs with relatedness data

#Mege genetic distance and geogrpahic distance
df_merged <- pairs_ibs %>%
  inner_join(dist2, by = "canon_pair")

#write.csv(df_merged, "genetic_geographic_distance_fulldataset.csv")
```

```{r}
#Preparing data - gen dist + geo dist - full dataset and filtered dataset

diff_cluster <- c("golden_2","golden_3","golden_21","golden_64",
               "grunt_159","grunt_181","long_209","long_219","grunt_127", "golden_52_d", "golden_55_d")

lin_map <- tibble(sample = samples,
                  lineage = if_else(sample %in% diff_cluster, "out","main")) #mapping to which cluster each sample blong to

gd<- df_merged %>%
  transmute(Individual1 = std_id(Individual1.x),
            Individual2 = std_id(Individual2.x),
            gen_dist, n_loci, Distance_m)

#all samples - full dataset
gd_all <- gd %>%
  left_join(lin_map %>% rename(Individual1 = sample, lin1 = lineage), by = "Individual1") %>%
  left_join(lin_map %>% rename(Individual2 = sample, lin2 = lineage), by = "Individual2") %>%
  mutate(distance_km = Distance_m/1000,
         log_dist    = log1p(distance_km),
         i = factor(Individual1),
         j = factor(Individual2),
         w_std = n_loci/mean(n_loci, na.rm=TRUE)) #weigth loci

#adding env metrics from sponge for each individual in the pair
gd_sponge_all <- gd_all %>%
  left_join(sponge_meta %>% 
              dplyr::select(id, sponge_id, S_width_cm, S_species, Reef) %>%
              rename(Individual1 = id, 
                     sp1 = sponge_id, 
                     size1 = S_width_cm,
                     species1 = S_species,
                     reef1 = Reef),
            by = "Individual1") %>%
  left_join(sponge_meta %>% 
              dplyr::select(id, sponge_id, S_width_cm, S_species, Reef) %>%
              rename(Individual2 = id, 
                     sp2 = sponge_id, 
                     size2 = S_width_cm,
                     species2 = S_species,
                     reef2 = Reef),
            by = "Individual2") %>%
  mutate(
    same_sponge      = sp1 == sp2,
    same_reef        = reef1 == reef2,
    same_species     = species1 == species2,
    diff_sponge_sz   = abs(size1 - size2),
    log_dist_km      = log1p(distance_km),
    same_reef        = factor(same_reef),
    same_sponge      = factor(same_sponge),
    same_species     = factor(same_species),
    i                = factor(Individual1),
    j                = factor(Individual2)) #creating variables combining each indv from dyad

#only main cluster - filtered dataset
gd_main <- gd %>%
  left_join(lin_map %>% rename(Individual1 = sample, lin1 = lineage), by = "Individual1") %>%
  left_join(lin_map %>% rename(Individual2 = sample, lin2 = lineage), by = "Individual2") %>%
  filter(lin1=="main", lin2=="main") %>%
  mutate(distance_km = Distance_m/1000,
         log_dist    = log1p(distance_km),
         i = factor(Individual1),
         j = factor(Individual2),
         w_std = n_loci/mean(n_loci, na.rm=TRUE)) 

#adding env metrics from sponge for each individual in the pair
gd_sponge_main <- gd_main %>%
  left_join(sponge_meta %>% 
              dplyr::select(id, sponge_id, S_width_cm, S_species, Reef) %>%
              rename(Individual1 = id, 
                     sp1 = sponge_id, 
                     size1 = S_width_cm,
                     species1 = S_species,
                     reef1 = Reef),
            by = "Individual1") %>%
  left_join(sponge_meta %>% 
              dplyr::select(id, sponge_id, S_width_cm, S_species, Reef) %>%
              rename(Individual2 = id, 
                     sp2 = sponge_id, 
                     size2 = S_width_cm,
                     species2 = S_species,
                     reef2 = Reef),
            by = "Individual2") %>%
  mutate(
    same_sponge      = sp1 == sp2,
    same_reef        = reef1 == reef2,
    same_species     = species1 == species2,
    diff_sponge_sz   = abs(size1 - size2),
    log_dist_km      = log1p(distance_km),
    same_reef        = factor(same_reef),
    same_sponge      = factor(same_sponge),
    same_species     = factor(same_species),
    i                = factor(Individual1),
    j                = factor(Individual2))#creating variables combining each indv from dyad

```

#### Full population-genomic dataset

This model uses all 42 biological individuals (861 unique dyads) to test whether genetic distance varies with geographic separation, shared sponge occupancy, difference in sponge size, shared reef, or shared host species. Random intercepts for both members of each dyad account for the repeated appearance of individuals across pairwise observations.

```{r}
#LMM: gen_dist ~ geo_dist + env metrics (full dataset: 42 indv = 861 dyads)
m_all<- lmer(
  gen_dist ~ log_dist_km + same_sponge + diff_sponge_sz + same_reef + same_species +
    (1 | i) + (1 | j),
  data = gd_sponge_all
)

summary(m_all)
```

```{r}
#Checking model fit 
model_performance(m_all)
check_model(m_all)
check_collinearity(m_all)
isSingular(m_all) #small but still present
r2(m_all) #fixed effects explain less than 1% of genetic distance
```

#### Filtered main-lineage dataset

The same model is repeated for the 33 individuals retained in the main-lineage dataset (528 unique dyads). Holding the model structure constant permits direct comparison with the complete dataset and tests whether the reported spatial and ecological patterns persist after removal of the documented high-F individuals and technical replicates. The model summary and diagnostic outputs support the filtered-dataset results visualized in Figure 2C-D.

```{r}
#LMM: gen_dist ~ geo_dist + env metrics (filtered dataset: 33 indv = 528 dyads)
m_main<- lmer(
  gen_dist ~ log_dist_km + same_sponge + diff_sponge_sz + same_reef + same_species +
    (1 | i) + (1 | j),
  data = gd_sponge_main
)

summary(m_main)
```

```{r}
#Checking model fit 
model_performance(m_main)
check_model(m_main)
check_collinearity(m_main)
isSingular(m_main) #small but still present
r2(m_main) #fixed effects explain only 8% of genetic distance
```

#### Visualize the filtered genetic-distance results

```{r}
#Figure 2C - Genetic-distance distribution

#arranging df
gd_sponge_main_v2 <- gd_sponge_main %>%
  mutate(
    distance_km = coalesce(distance_km, Distance_m / 1000),
    dist_bin = cut(
      distance_km,
      breaks = c(0, 0.025, 0.075, 0.2, 3.5, Inf),
      labels = c("≤25 m", "25–75 m", "75–200 m", "200–3500 m", ">3500 m"),
      include.lowest = TRUE),
    dyad_loc = case_when(
      as.logical(same_sponge) ~ "same group",
      as.logical(same_reef)   ~ "same reef",
      TRUE                    ~ "different reefs"),
    dyad_loc = factor(dyad_loc,
                      levels = c("different reefs", "same reef", "same group")))

means_df_gd <- gd_sponge_main_v2 %>%
  dplyr::group_by(dyad_loc) %>%
  dplyr::summarise(
    mean_gd = mean(gen_dist, na.rm = TRUE),
    .groups = "drop"
  )

#plotting
cols_gd <- c(
  "same group"       = "grey70",
  "same reef"       = "darkseagreen3",
  "different reefs" = "darkslategray4"
)

ggplot(gd_sponge_main_v2, aes(x = gen_dist, fill = dyad_loc)) +
  geom_density(alpha = 0.7, adjust = 1.0, color = NA) +
  geom_vline(
    data = means_df_gd,
    aes(xintercept = mean_gd, colour = dyad_loc),
    linewidth = 0.7,
    linetype = "dashed",
    show.legend = FALSE
  ) +
  scale_fill_manual(
    values = cols_gd,
    name   = "Dyad location"
  ) +
  scale_colour_manual(
    values = cols_gd
  ) +
  labs(
    x = "Genetic distance (1 - IBS)",
    y = "Density"
  ) +
  theme_classic() +
  theme(
    legend.position      = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background    = element_rect(fill = "white", colour = NA)
  )

ggsave(file.path(output_dir, "Fig_2C.pdf"), width = 5, height = 4, units = "in")

```

```{r}
#Figure 2D - Genetic distance by distance class

#arranging df
bin_summ_gd<- gd_sponge_main_v2 %>%
  group_by(dist_bin) %>%
  summarise(
    n_dyads = n(),
    mean_gd = mean(gen_dist, na.rm = TRUE),
    se_gd   = sd(gen_dist, na.rm = TRUE) / sqrt(n()),
    .groups   = "drop"
  )

#plotting
ggplot(gd_sponge_main_v2, aes(x = dist_bin, y = gen_dist)) +
  # raw points
  geom_jitter(
    width  = 0.15,
    size   = 2,
    alpha  = 0.35,
    color  = "aquamarine4"
  ) +
  # bin means
  geom_point(
    data        = bin_summ_gd,
    aes(x = dist_bin, y = mean_gd),
    inherit.aes = FALSE,
    size        = 3,
    color       = "black",
    fill        = "black",
    shape       = 21,
    stroke      = 0.8
  ) +
  # error bars
  geom_errorbar(
    data        = bin_summ_gd,
    aes(x = dist_bin,
        ymin = mean_gd - se_gd,
        ymax = mean_gd + se_gd),
    inherit.aes = FALSE,
    width       = 0.15,
    color       = "black"
  ) +
  labs(
    x = "Dyad distance category",
    y = "Genetic distance (1 - IBS)"
  ) +
  theme_classic() +
  theme(
    axis.text.x      = element_text(size = 10),
    axis.text.y      = element_text(size = 10),
    strip.background = element_blank(),
    legend.position  = "none"
  )

ggsave(file.path(output_dir, "Fig_2D.pdf"), width = 5, height = 4, units = "in")
```

## 3. Population-genomic summaries and PCA

The population VCF is converted to `genind` and GDS representations to calculate heterozygosity and inbreeding. Figure S6 presents the PCA of 41 biological individuals: the two technical replicates and the PCA outlier `grunt_159` are excluded, while the remaining documented high-F individuals are retained. Figure S7 provides the corresponding reduced-dataset sensitivity comparison after all documented high-F individuals are removed.

```{r}
#Converting SNPs to genind object

#Full dataset
vcf_full  <- read.vcfR(vcf_all)
genind_full <- vcfR2genind(vcf_full) #FULL DATASET
pop(genind_full) <- factor(rep("pop1", nInd(genind_full))) #same population
hf_full <- genind2hierfstat(genind_full)

#Fitered dataset
# individuals to remove
rm_inds <- c("golden_2",  "golden_3",  "golden_21", "golden_64", "grunt_127", "grunt_159" ,"grunt_181", "long_209",  "long_219" , "golden_52_d", "golden_55_d")

genind_fil <- genind_full[!indNames(genind_full) %in% rm_inds, ]
hf_fil <- genind2hierfstat(genind_fil)

```


```{r}
basic_stats_all <- basic.stats(hf_full)
basic_stats_all$overall["Fis"] #Overall (global) inbreeding coefficient F_IS 0.363

basic_stats_fil <- basic.stats(hf_fil) #filtered dataset
basic_stats_fil$overall["Fis"] 

Ho_all <- basic_stats_all$Ho # Observed heterozygosity per locus
He_all <- basic_stats_all$Hs # Expected heterozygosity per locus

Ho_fil <- basic_stats_fil$Ho 
He_fil <- basic_stats_fil$Hs 

#mean overall
mean(Ho_all, na.rm = TRUE) #observed 0.2115303
mean(He_all, na.rm = TRUE) #expected 0.3320372

mean(Ho_fil, na.rm = TRUE) #observed 0.2284386
mean(He_fil, na.rm = TRUE) #expected 0.2609608

head(basic_stats_all$overall)

```

```{r}
#Using SNPRelate package
all_ids <- read.gdsn(index.gdsn(genofile_all, "sample.id")) #get all samples id

# Figure S6 uses the biological genotyped dataset after excluding the PCA
# outlier grunt_159. The two technical replicates are also excluded; the
# remaining documented high-F biological individuals are retained.
remove_ids <- c("golden_52_d", "golden_55_d", "grunt_159")

keep_ids <- setdiff(all_ids, remove_ids)

#running PCA
pca_all <- snpgdsPCA(
  genofile_all,
  sample.id = keep_ids,
  autosome.only = TRUE
)

#generate PCA df
pca.df.all<- data.frame(
  id = pca_all$sample.id,
  PC1 = pca_all$eigenvect[,1],
  PC2 = pca_all$eigenvect[,2],
  PC3 = pca_all$eigenvect[,3]
)
#Merge PCA df with metadata
pca.df.all<- merge(pca.df.all, meta_all, by = "id")
head(pca.df.all)

```

```{r}
#Plotting PCA
pc.percent <- pca_all$varprop * 100 #obtain percent variance explained

PCA_all<- ggplot(pca.df.all, aes(PC1, PC2, color = site, label = id)) +
  geom_point(size = 2) +
  ggrepel::geom_text_repel(
    size = 3,
    max.overlaps = Inf, 
    box.padding  = .4,
    point.padding = .25,
    min.segment.length = 0
  ) +
  labs(
    x = paste0("PC1 (", round(pc.percent[1], 1), "%)"),
    y = paste0("PC2 (", round(pc.percent[2], 1), "%)")
  ) +
 scale_color_manual(values = c("curlew"="#008B8B", "golden" = "#DAA520", "grunt" = "#6B8E23", "long" = "#000080")) +
  theme_bw()

PCA_all

ggsave(file.path(output_dir, "Figure_S6.pdf"), width = 8, height = 6, units = "in")

```
PC1 is explaining a lot - which means a strong separation between the two clusters. This is probably driven by the little cluster being really similar to each other - very homozygous.

### Missing-data sensitivity check

Sample-level missingness is compared with PCA position to verify that clustering is not an artifact of uneven genotype completeness.
```{r}
miss <- snpgdsSampMissRate(genofile_all)
sample.id <- read.gdsn(index.gdsn(genofile_all, "sample.id"))

miss.df <- data.frame(
  id = sample.id,
  miss = miss
) 
head(miss.df) #less than 10% missing data in all samples
```

```{r}

miss_threshold <- 0.3

bad_ids <- miss.df$id[miss.df$miss > miss_threshold]

bad_ids
bad_ids <- c( "golden_2",  "golden_3",  "golden_21", "golden_64", "grunt_127", "grunt_159" ,"grunt_181", "long_209",  "long_219" , "golden_52_d", "golden_55_d") #samples with highest missing data - let's see if removing them would change the PC structure

keep_ids <- setdiff(miss.df$id, bad_ids)

pca_test <- snpgdsPCA(genofile_all, sample.id = keep_ids)

pca.df.test<- data.frame(
  id = pca_test$sample.id,
  PC1 = pca_test$eigenvect[,1],
  PC2 = pca_test$eigenvect[,2],
  PC3 = pca_test$eigenvect[,3]
)

pca.df.test<- merge(pca.df.test, meta_all, by = "id")
head(pca.df.test)
```

```{r}
pc.percent.test <- pca_test$varprop * 100

pc.percent.test[1:3]

PCA_all.test<- ggplot(pca.df.test, aes(PC1, PC2, color = site, label = id)) +
  geom_point(size = 2) +
  ggrepel::geom_text_repel(
    size = 3,
    max.overlaps = Inf, 
    box.padding  = .4,
    point.padding = .25,
    min.segment.length = 0
  ) +
  labs(
    x = paste0("PC1 (", round(pc.percent[1], 1), "%)"),
    y = paste0("PC2 (", round(pc.percent[2], 1), "%)")
  ) +
 scale_color_manual(values = c("curlew"="#008B8B", "golden" = "#DAA520", "grunt" = "#6B8E23", "long" = "#000080")) +
  theme_bw()

PCA_all.test 

ggsave(file.path(output_dir, "Figure_S7.pdf"), width = 8, height = 6, units = "in")
```
Missing data is low across samples and therefore is not driving clustering separation.

## Session information

The following output records the R, operating-system, and package versions used to render this report.

```{r session-information}
sessionInfo()
```
