---
title: "Risor ruber - Social and life history analyses"
date: "2026-08-12"
---
This code refers to Bunholi et al. 2026 "Ecological constraints of habitat specialization shape the social organization and life-history of a sponge-dwelling goby."

## Analysis overview

This report reproduces the sponge-occupancy, social-organization, reproductive, and age-structure analyses.
Manuscript figures: 1C, 3, 4, and 5.
Supplementary figures: S5, S8, S9, S10, S11, and S12.

### Required inputs

Both inputs are provided in `data/social_life_history/`:

- "sponge_occupancy_df.csv"

- "social_data_all_complete.csv"

User: please run this R Markdown file from the Scripts/ directory.

### Software setup

The following chunk lists the packages required by this report.

```{r setup-packages}
# Install required packages before the first run.
# Uncomment this command only if the packages are not already installed.
# install.packages(c(
#   "tidyverse", "lme4", "lmerTest", "performance", "ggeffects",
#   "piecewiseSEM", "ggdist", "ggthemes", "emmeans", "AICcmodavg",
#   "glmmTMB", "scales", "patchwork"
# ))

# Load the packages used in this analysis.
pkgs <- c(
  "tidyverse", "lme4", "lmerTest", "performance", "ggeffects",
  "piecewiseSEM", "ggdist", "ggthemes", "emmeans", "AICcmodavg",
  "glmmTMB", "scales", "patchwork"
)
invisible(lapply(pkgs, require, character.only = TRUE))
```

## 1. Sponge occupancy

This section tests whether sponge width predicts occupancy by *Risor ruber* and it generated Figure 1C. Counts and proportions are calculated by reef to document spatial variation in the availability of occupied sponges.
```{r}
data_dir <- file.path("..", "data", "social_life_history")
output_dir <- file.path("..", "outputs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
social_occ <- read.csv(file.path(data_dir, "sponge_occupancy_df.csv"))

social_occ$Sponge_ID <- as.factor(social_occ$Sponge_ID)
social_occ$R_ruber <- as.character(social_occ$R_ruber)
social_occ$Reef <- as.factor(social_occ$Reef)

social_occ<- social_occ %>%
  mutate(z_Max_width_cm = scale(Max_width_cm))

#hist(social_occ$Max_width_cm)
```

```{r}
social_occ_2<- social_occ %>%
  filter(R_ruber == "Present")
summary(social_occ_2)
```

#### Site-level occupancy summary

```{r}
#summarizing per site

site_occupancy <- social_occ %>%
  filter(!Reef == "Xesto") %>%
  group_by(Reef) %>%
  summarise(
    n_total     = n(),
    n_occupied  = sum(S_occ == 1),
    prop_occupied = n_occupied / n_total,
    .groups = "drop"
  )

site_occupancy
```

```{r}
site_occupancy %>%
  summarise(
    min_prop = min(prop_occupied),
    max_prop = max(prop_occupied)
  )
site_occupancy %>%
  mutate(percent_occupied = prop_occupied * 100)

```

#### MODEL - glmer with binomial distribution (presence/absence)

```{r}
glmer.occ <- glmer(cbind(S_occ,S_unocc)~z_Max_width_cm+(1|Reef), family = binomial(link="logit"), data = social_occ)

summary(glmer.occ) 

r2(glmer.occ)

```

Interpretation: *Risor ruber* occupancy increased significantly with sponge size. In a binomial GLMM with reef as a random intercept, maximum sponge width had a positive effect on the probability of occupancy (β = 0.10 ± 0.02 SE, z = 4.42, P < 0.001), corresponding to roughly an 11% increase in the chances of occupation for each additional centimeter of sponge width. The model explained 11.9% of the variance via fixed effects (marginal R² = 0.119) and 16.9% when accounting for among-reef differences (conditional R² = 0.169).

```{r}
#Model checks and metrics
check_model(glmer.occ) 
check_overdispersion(glmer.occ)
check_collinearity(glmer.occ)
r2(glmer.occ) 

sd_width <- sd(social_occ$Max_width_cm, na.rm = TRUE) #6.760123

OR_1SD <- exp(0.6869) #1.987545
OR_1cm <- exp(0.6869 / sd_width) #1.106952
```

Predictions from the standardized-width model are returned to centimeters before being plotted with the observed occupancy data. The threshold annotations follow the observed prediction grid.

```{r}
#Plot Model - Sponge occupancy vs sponge size + Raw data 
occ.plot <- ggpredict(glmer.occ,
                      terms = c("z_Max_width_cm [all]"),
                      type = "random")
occ_df <- as.data.frame(occ.plot)

width_mean <- mean(social_occ$Max_width_cm, na.rm = TRUE)
width_sd <- sd(social_occ$Max_width_cm, na.rm = TRUE)
occ_df$x <- occ_df$x * width_sd + width_mean

#Find the observed-range prediction-grid width closest to each probability.
get_size_from_grid <- function(p, df = occ_df) {
  df$x[which.min(abs(df$predicted - p))]
}

probs <- c(0.01, 0.5, 0.85)

thresholds <- data.frame(
  prob  = probs,
  size  = sapply(probs, get_size_from_grid),
  label = c("1%", "50%", "85%"))

#Extracting empirical info
min_present <- min(social_occ$Max_width_cm[social_occ$S_occ == 1], na.rm = TRUE)

empirical_df <- data.frame(
  size  = min_present,
  label = "Min occupied")

social_occ$S_occ_f <- factor(social_occ$S_occ,
                             levels = c(0, 1),
                             labels = c("Absent", "Present"))

#Plotting the model
pred.plot.occ <-
  ggplot() +
  geom_ribbon(data = occ_df,
              aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2) +
  geom_line(data = occ_df,
            aes(x = x, y = predicted),
            size = 1) +
  geom_point(data = social_occ,
    aes(x = Max_width_cm, y = S_occ, colour = S_occ_f),
    position = position_jitter(width = 0.3, height = 0),
    alpha = 0.7,
    size = 2)+
  #adding 1%, 50%, 90% thresholds
  geom_vline(data = thresholds,
             aes(xintercept = size),
             linetype = "dashed") +
  geom_text(data = thresholds,
            aes(x = size,y = 0.15, label = paste0(label, " (", round(size, 1), " cm)")),
            angle = 90, vjust = -0.4, size = 3) +
  #adding empirical min occupied sponge
  geom_vline(data = empirical_df,
             aes(xintercept = size), linetype = "solid") +
  geom_text(data = empirical_df,
            aes(x = size, y = 0.20, label = paste0("Min occupied (", round(size, 1), " cm)")),
            angle = 90, vjust = -0.4, size = 3) +
  # aesthetics
  scale_colour_manual(values = c("Absent" = "grey60", "Present" = "black")) +
  scale_y_continuous("Sponge occupancy",
                     limits = c(0, 1),
                     breaks = c(0, 0.25, 0.5, 0.75, 1)) +
  scale_x_continuous("Maximum Sponge Width (cm)") +
  theme_bw(base_size = 14) +
  theme(legend.position = "none",
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "black"),
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.background = element_blank())

pred.plot.occ

ggsave(file.path(output_dir, "Figure_1C.pdf"), width = 6, height = 5, units = "in")

```

```{r}
#Complementary density plot
social_occ %>%
  ggplot(aes(x = factor(R_ruber), y = Max_width_cm, fill = factor(R_ruber)))+
  theme_bw() +
  theme(legend.position = "none",
        axis.text = element_text(color = "black", size = 10),
        axis.title = element_text(size = 15),
        panel.background = element_blank(),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank())+
  stat_halfeye(
    adjust = 0.4,
    justification = 0,
    .width = 0,
    point_colour = NA
  )  +
  stat_dots(
    side = "left",
    justification = 1,
    binwidth = 0.3)+
  scale_fill_tableau("Tableau 20", name = NULL) +
  labs(
    x = "Sponge occupancy",
    y = "Maximum Sponge Width (cm)",
    fill = "Cylinders"
  ) +
  coord_flip()
ggsave(file.path(output_dir, "Figure_1C.2.pdf"), width = 6, height = 5, units = "in")
```
Further refinement of Figure 1C was done in illustrator.

## 2. Descriptive social and habitat summaries

This section summarizes sampling by reef and sponge species, fish size, group size, sex, and social rank. These outputs document sample sizes and provide the descriptive results used in the text and supplementary figures.

Computing some basic descriptive metrics with the main social dataset before further analyses:

```{r}
social.all <- read.csv(file.path(data_dir, "social_data_all_complete.csv"))

head(social.all)
```

```{r}
social.all <- social.all %>%
  mutate(
    Reef      = factor(Reef),
    Sponge_id = factor(Sponge_id),
    Rank_n    = as.integer(as.character(Rank)),
    sex = case_when(
      str_to_lower(is_male) %in% c("male","m","1","true","t")        ~ "male",
      str_to_lower(is_male) %in% c("female","f","0","false","fa")    ~ "female",
      TRUE                                                           ~ NA_character_
    ) %>% factor(levels = c("female","male"))
  )
head(social.all)


social.all <- social.all %>%
  mutate(
    z_Sponge_width_cm = scale(Sponge_width_cm),
    z_TL_mm           = scale(TL_mm)
  )


```

#### Sponge metrics

```{r}
sponges_per_site <- social.all %>%
  distinct(Reef, Sponge_id) %>%      #each sponge counted once per reef
  count(Reef, name = "n_sponges")

sponges_per_site
```

```{r}
sponges_summary <- sponges_per_site %>%
  summarise(
    mean_sponges = mean(n_sponges),
    sd_sponges   = sd(n_sponges),
    min_sponges  = min(n_sponges),
    max_sponges  = max(n_sponges),
    n_sites = n()
  )

sponges_summary
```

```{r}
sponges_per_site_species <- social.all %>%
  distinct(Reef, S_species, Sponge_id) %>%  # each sponge once per reef × species
  count(Reef, S_species, name = "n_sponges")

sponges_per_site_species
```

```{r}
species_summary <- sponges_per_site_species %>%
  group_by(S_species) %>%
  summarise(
    n_occupied = sum(n_sponges),   #total occupied sponges of that species
    .groups = "drop"
  ) %>%
  mutate(
    percent_occupied = 100 * n_occupied / sum(n_occupied)
  )
species_summary <- species_summary %>%
  arrange(desc(percent_occupied))

```

#### Fish total length summaries

```{r}
#Overall
tl_overall <- social.all %>%
  summarise(
    n          = sum(!is.na(TL_mm)),
    mean_TL    = mean(TL_mm, na.rm = TRUE),
    sd_TL      = sd(TL_mm, na.rm = TRUE),
    min_TL     = min(TL_mm, na.rm = TRUE),
    max_TL     = max(TL_mm, na.rm = TRUE)
  )

tl_overall
```

```{r}
#By sex
tl_by_sex <- social.all %>%
  filter(!is.na(sex)) %>%
  group_by(sex) %>%
  summarise(
    n          = n(),
    mean_TL    = mean(TL_mm, na.rm = TRUE),
    sd_TL      = sd(TL_mm, na.rm = TRUE),
    min_TL     = min(TL_mm, na.rm = TRUE),
    max_TL     = max(TL_mm, na.rm = TRUE),
    .groups    = "drop"
  )

tl_by_sex
```

#### Group size distribution

```{r}
group_df <- social.all %>%
  distinct(Sponge_id, Reef, Group_size)

group_size_summary <- group_df %>%
  summarise(
    n_groups      = n(),
    median_gsize    = median(Group_size, na.rm = TRUE),
    min_gsize     = min(Group_size, na.rm = TRUE),
    max_gsize     = max(Group_size, na.rm = TRUE)
  )

group_size_summary
```

```{r}
group_size_freq <- group_df %>%
  count(Group_size, name = "n_groups") %>%
  arrange(Group_size)

group_size_freq
```

#### Rank distribution

```{r}
rank_counts <- social.all %>%
  count(Rank_n, name = "n_ind") %>%
  arrange(Rank_n)

rank_counts
```

#### Sex ratio overall and by rank

```{r}
#Overall sex ratio
sex_overall <- social.all%>%
  filter(!is.na(sex)) %>%
  count(sex, name = "n") %>%
  mutate(
    prop = n / sum(n)
  )

sex_overall
```

```{r}
# Sex by rank (for rank × sex figure and text)
sex_by_rank <- social.all %>%
  filter(!is.na(sex), !is.na(Rank_n)) %>%
  count(Rank_n, sex, name = "n") %>%
  group_by(Rank_n) %>%
  mutate(
    total_rank = sum(n),
    prop       = n / total_rank
  ) %>%
  ungroup() %>%
  arrange(Rank_n, sex)

sex_by_rank
```

#### Occupied sponges by reef and host species 

Occupied sponges are counted by reef and host species. The expected output is Supplementary Figure S5.

```{r}

sponge <- social.all %>%           # or your full social dataframe
  group_by(Sponge_id, S_species, Reef) %>%
  summarise(
    Group_size      = n(),
    Sponge_width_cm = first(Sponge_width_cm),
    .groups = "drop"
  ) #106 sponges


```

```{r}
counts <- sponge %>%
  count(Reef, S_species, name = "freq")

counts <- counts %>%
  group_by(Reef) %>%
  mutate(total_site = sum(freq)) %>%
  ungroup() %>%
  mutate(Reef = reorder(Reef, total_site))

counts
```

```{r}
ggplot(counts, aes(x = Reef, y = freq, fill = S_species)) +
  geom_col(position = "dodge", alpha = 0.9) +
  scale_fill_manual(values = c(
    "Ircinia campana"    = "rosybrown",
    "Ircinia felix"      = "saddlebrown",
    "Ircinia strobilina" = "grey28",
    "Ircinia ruetzleri"  = "peachpuff3" 
  )) +
  labs(
    x = "Sampling reef",
    y = "Frequency of Occupation by Risor ruber",
    fill = "Host sponge"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "top") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Figure_S5.pdf"), width = 7, height = 5, units = "in")

```

## 3. Social organization

This section quantifies adjacent-rank size ratios and tests whether group-size structure changes with group size and habitat characteristics. It produces the group-structure summaries and figures used to describe size-based social organization.

### Adjacent-rank size ratios

For each social group, the total length of a subordinate is divided by the length of the individual immediately above it in rank. Ratios closer to one indicate more similar body sizes, whereas smaller ratios indicate a steeper size hierarchy. The expected output is Supplementary Figure S8.

```{r}
#orderining individuals within each group
df_ratios<- social.all %>%
  mutate(Rank_n = as.numeric(as.character(Rank)))%>%
  arrange(Sponge_id, Rank_n)

# add "next" fish in rank and compute adjacent size ratio
df_ratios <- df_ratios %>%
  group_by(Sponge_id) %>%
  arrange(Rank_n, .by_group = TRUE) %>%
  mutate(
    next_TL  = lead(TL_mm),
    next_rank = lead(Rank_n),
    ratio    = next_TL / TL_mm,
    Ratio_ID = ifelse(
      is.na(next_TL),
      NA_character_,
      sprintf("R%02d/R%02d", next_rank, Rank_n)
    )
  ) %>%
  ungroup()

#summarize to group-level means (Group.ratio)
group_ratio_df <- df_ratios %>%
  group_by(Sponge_id, Reef) %>%                  
  summarise(
    Group.size  = n(),                            
    Group.ratio = if (sum(!is.na(ratio)) > 0) {
      mean(ratio, na.rm = TRUE)                 
    } else {
      NA_real_                                
    },
    .groups = "drop"
  )


#transforming to wide format
ratio_wide <- df_ratios %>%
  filter(!is.na(ratio)) %>%                
  dplyr::select(Sponge_id, Ratio_ID, ratio) %>%
  pivot_wider(
    names_from  = Ratio_ID,
    values_from = ratio
  )

summary_wide <- group_ratio_df %>%
  left_join(ratio_wide, by = "Sponge_id")


#group ratio - long format
ratio_long_df <- df_ratios %>%
  filter(!is.na(ratio)) %>%
  transmute(
    Sponge_id,
    Reef,
    Group.size = ave(Rank_n, Sponge_id, FUN = length), 
    Ratio_ID,
    Ratio_long = ratio
  )

ratio_long_df <- df_ratios %>%
  mutate(Rank_n = as.numeric(as.character(Rank))) %>%
  arrange(Sponge_id, Rank_n) %>%
  group_by(Sponge_id) %>%
  mutate(
    next_TL   = dplyr::lead(TL_mm),
    next_rank = dplyr::lead(Rank_n),
    Ratio_long = next_TL / TL_mm,
    Ratio_ID   = ifelse(
      is.na(next_TL),
      NA_character_,
      sprintf("R%02d/R%02d", next_rank, Rank_n)
    )
  ) %>%
  ungroup() %>%
  filter(!is.na(Ratio_long), !is.na(Ratio_ID))

```

```{r}
#obtain ratio summary of each individual
ratio_summary <- ratio_long_df %>%
  group_by(Ratio_ID) %>%
  summarise(
    mean_ratio = mean(Ratio_long, na.rm = TRUE),
    sd_ratio   = sd(Ratio_long, na.rm = TRUE),
    se_ratio   = sd_ratio / sqrt(n()),
    n          = n(),
    .groups    = "drop"
  ) %>%
  mutate(Ratio_ID = factor(Ratio_ID, levels = sort(unique(Ratio_ID))))

ratio_summary
```

The following plot retains rank pairs with sufficient replication to calculate a standard error.

```{r}
# reorder rank pairs in natural order
summary_ratios <- ratio_summary %>%
  mutate(Ratio_ID = factor(Ratio_ID, levels = Ratio_ID))%>%
  filter(!is.na(se_ratio))

ratio_long_clean <- ratio_long_df %>%
  filter(Ratio_ID %in% summary_ratios$Ratio_ID)

ggplot() +
  # Raw ratios (transparent points)
  geom_jitter(data = ratio_long_clean,
              aes(x = Ratio_ID, y = ratio, color = Reef),
              width = 0.1, height = 0,
              alpha = 0.4, size = 2) +
  
  # Means + SE
  geom_point(data = summary_ratios,
             aes(x = Ratio_ID, y = mean_ratio),
             size = 4, color = "black") +
  
  geom_errorbar(data = summary_ratios,
                aes(x = Ratio_ID,
                    ymin = mean_ratio - se_ratio,
                    ymax = mean_ratio + se_ratio),
                width = 0.1, linewidth = 0.7) +
  scale_color_manual(values = c("CUR"="#008B8B", "GOL" = "#DAA520", "GRU" = "#6B8E23", "LON" = "#000080"))+
  labs(x = "Rank pair",
       y = "Mean Size Ratio (+/- SE)") +
  
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave(file.path(output_dir, "Figure_S8.pdf"), width = 7, height = 5, units = "in")

```

### Group size and hierarchy steepness

The beta mixed model tests whether adjacent-rank size ratios change with group size while accounting for reef. The model evaluates whether larger groups have steeper or flatter size hierarchies.

Determine how steep or shallow the size hierarchy is
Mean ratio close to 1 - means individuals are very similar in size (weak hierarchy)
Mean ratio lower than that - individuals differ strongly - steep hierarchy

This number is a shape descriptor of the dominance/size structure of a group

Model - when groups get larger, are size differences between rank positions amplified or compressed? Does adding more individuals reinforce the hierarchy, or dilute it?

H1: Larger groups -> steeper hierarchy (lower ratios)
This would suggest strong competition within confined space; bigger individual monopolize resources; subordinate individuals get smaller relative to dominants

H2: Larger groups -> flatter hierarchy (higher ratios)
This would suggest that groups tolerate similarly sized individuals; size-based dominance is relaxed; the resource (sponge) provided enough microhabitats. 
This is probably the most plausible scenario for microhabitat-limited species


```{r}
#Model - does the size structure of groups depend on how large the group is?
#Do bigger groups have more competitive, more size-structured hierarchies?

m_gratio = glmmTMB(Group.ratio ~ Group.size + (1|Reef), family = beta_family(link = "logit"), data=summary_wide)

summary(m_gratio)

```

```{r}
#Model checks
check_model(m_gratio)
check_residuals(m_gratio) #uniformly distributed
model_performance(m_gratio)
```

#### Visualizing social structure across group sizes

```{r}
social_data<- social.all

#number of R. ruber located within one sponge ranged from 1-14 individuals 
#N of each group size: 1=50, 2=30, 3=11, 4=7, 5=3, 7=2, 10=1, 14=1). 
head(social_data)

str(social_data)

#change string
social_data$Group_size <- as.integer(social_data$Group_size)
social_data$ind_id <- as.factor(social_data$ind_id)
social_data$Rank <- as.factor(social_data$Rank)
social_data$Sponge_id <- as.factor(social_data$Sponge_id)
social_data$S_species <- as.factor(social_data$S_species)
social_data$Sponge_id <- as.factor(social_data$Sponge_id)
social_data$S_colour <- as.factor(social_data$S_colour)
social_data$Reef <- as.factor(social_data$Reef)

#summarize data
sumdat <- social_data %>%
  group_by(Rank, Group_size) %>%
  summarise(
    meanTL = mean(TL_mm, na.rm = TRUE),
    seTL   = sd(TL_mm, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(Rank = as.integer(as.character(Rank)))

keep <- c(1,2,3,4,5,7,10,14)

sumdat <- sumdat %>%
  filter(Group_size %in% keep)

```

```{r}
#Organize sclaes
ymin <- min(sumdat$meanTL - sumdat$seTL, na.rm = TRUE)
ymax <- max(sumdat$meanTL + sumdat$seTL, na.rm = TRUE)

#base theme
base_theme <- theme_bw() +
  theme(
    legend.position = "none",
    panel.grid = element_blank(),
    axis.title.y = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank()
  )

#individual panels
p1 <- ggplot(filter(sumdat, Group_size==1),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,4), breaks=1:4) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  labs(y="Mean TL") +
  theme_bw() +
  theme(panel.grid = element_blank())

p2 <- ggplot(filter(sumdat, Group_size==2),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,4), breaks=1:4) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p3 <- ggplot(filter(sumdat, Group_size==3),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,4), breaks=1:4) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p4 <- ggplot(filter(sumdat, Group_size==4),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,4), breaks=1:4) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p5 <- ggplot(filter(sumdat, Group_size==5),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,5), breaks=1:5) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p7 <- ggplot(filter(sumdat, Group_size==7),
             aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,7), breaks=1:7) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p10 <- ggplot(filter(sumdat, Group_size==10),
              aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,10), breaks=1:10) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

p14 <- ggplot(filter(sumdat, Group_size==14),
              aes(Rank, meanTL)) +
  geom_line(color="#34446e") +
  geom_point(color="#34446e", size=2) +
  geom_linerange(aes(ymin=meanTL-seTL, ymax=meanTL+seTL),
                 color="#34446e") +
  scale_x_continuous(limits=c(1,14), breaks=1:14) +
  scale_y_continuous(limits=c(floor(ymin), ceiling(ymax))) +
  base_theme

#combine with patchwork
pall<- (p1 | p2 | p3 | p4 | p5 | p7 | p10 | p14) +
  plot_layout(widths = c(3, 3, 3, 3, 3, 5, 5, 8))

pall

ggsave(file.path(output_dir, "Figure_3A.pdf"), plot = pall, width = 12, height = 4, dpi = 600)
```
Further refinement of Figure 3A was done in illustrator.

### Piecewise structural equation model

The piecewise SEM evaluates the proposed path from sponge width to dominant size, sequential subordinate sizes, and group size. The section reports component-model summaries, global fit, and the directed graph.

```{r}
#Organizing df for SEM model
sem_df <- social.all %>%
  mutate(Rank_n  = as.integer(Rank),
         rank_lab = paste0("R", Rank_n, "_TL")) %>%
  dplyr::filter(Rank_n %in% 1:3) %>%
  dplyr::select(Sponge_id,Reef, S_width_cm = Sponge_width_cm, Group_size, rank_lab, TL_mm) %>%
  pivot_wider(names_from  = rank_lab, values_from = TL_mm) %>%
  filter(!is.na(R1_TL), !is.na(R2_TL), !is.na(R3_TL))

```

```{r}
#tidy data and check distribution of target variable
sem_df <- sem_df %>%
  mutate(log_width = log(S_width_cm),
         Reef = factor(Reef))

#hist(social.data.clean$Group_size) #poisson distribution
```

```{r, message= FALSE}
#SEM - Piecewise structural equation modeling
#unites structural equations in a single model

sem_mod <- psem(
  lmer(R1_TL ~ log_width + (1|Reef), data = sem_df),
  lmer(R2_TL ~ R1_TL     + (1|Reef), data = sem_df),
  lmer(R3_TL ~ R2_TL     + (1|Reef), data = sem_df),
  glmer(Group_size ~ log_width + R1_TL + (1|Reef),
        family = poisson, data = sem_df)
)

sem_output <- summary(sem_mod)

```

```{r}
plot(sem_mod) #the actual paper figure (Figure 3B) was produced on illustrator.
```

Interpretation: 

SEM - Network of linked responses, including direct and indirect effects.
The relationships are hierarchical / causal (rank 1 → rank 2 → rank 3), not just correlations.
The global goodness-of-fit tests. High P-values (> 0.05) mean: The hypothesized network of causal paths is not rejected by the data – it’s statistically consistent with the observed covariances.
 
### Host species and group size

The following negative-binomial models test whether host species adds explanatory value beyond sponge width for variation in group size. Only host species represented by at least 10 sponges are retained. Model comparison and the species-specific prediction curves provide the results shown in Supplementary Figure S9.

```{r}
#Restricting to only abundant species (>10) before modeling 

sponge <- social.all %>%          
  group_by(Sponge_id, S_species, Reef) %>%
  summarise(
    Group_size      = n(),
    Sponge_width_cm = first(Sponge_width_cm),
    .groups = "drop"
  ) #106 sponges


species_counts <- table(sponge$S_species)

common_species <- names(species_counts[species_counts >= 10])

common_species
```

```{r}
groups_main <- sponge %>%
  filter(S_species %in% common_species)

m_mix_width_main <- glmmTMB(
  Group_size ~ scale(Sponge_width_cm) + (1 | Reef),
  family = nbinom2,
  data   = groups_main
)

m_mix_main <- glmmTMB(
  Group_size ~ scale(Sponge_width_cm) + S_species + (1 | Reef),
  family = nbinom2,
  data   = groups_main
)

AIC(m_mix_width_main, m_mix_main)

```

```{r}
summary(m_mix_main)
```

Group size increased strongly with sponge width, and host identity accounted for additional variation beyond this effect (negative binomial GLMM, three common host species). For sponges of the same width, groups in I. strobilina were substantially larger than those in I. campana (≈ exp(0.81) ≈ 2.2× larger), whereas groups in I. felix did not differ significantly from I. campana.

```{r}
pred_group <- ggpredict(m_mix_main, terms = c("Sponge_width_cm [all]", "S_species"))
```

```{r}
ggplot() +
  geom_ribbon(
    data = pred_group,
    aes(x = x, ymin = conf.low, ymax = conf.high, fill = group),
    alpha = 0.15
  ) +
  geom_line(
    data = pred_group,
    aes(x = x, y = predicted, colour = group),
    linewidth = 1.1
  ) +
  geom_point(
    data = groups_main,
    aes(x = Sponge_width_cm, y = Group_size, colour = S_species),
    size = 2, alpha = 0.7
  ) +
  scale_color_manual(values = c(
    "Ircinia campana"    = "rosybrown",
    "Ircinia felix"      = "saddlebrown",
    "Ircinia strobilina" = "grey28"
  )) +
  scale_fill_manual(values = c(
    "Ircinia campana"    = "rosybrown",
    "Ircinia felix"      = "saddlebrown",
    "Ircinia strobilina" = "grey28"
  ), guide = "none") +
  labs(
    x = "Sponge width (cm)",
    y = "Predicted group size",
    colour = "Host sponge"
  ) +
  theme_classic(base_size = 14) +
  theme(legend.position = "top")

ggsave(file.path(output_dir, "Figure_S9.pdf"), width = 7, height = 5, units = "in")

```

## 4. Life history

This section evaluates sex and rank patterns, reproductive status and egg number, and the relationship between otolith-derived age and body size.

### Sexual structure - Figure 4A and 4B

```{r}
#social data
d <- social.all %>%
  mutate(
    Reef      = factor(Reef),
    Sponge_id = factor(Sponge_id),
    Rank      = factor(Rank),
    # robust sex mapping (edit levels if needed)
    sex = case_when(
      str_to_lower(is_male) %in% c("male","m","1","true","t") ~ "male",
      str_to_lower(is_male) %in% c("female","f","0","false","fa") ~ "female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("female","male")),
    # convenience scalers
    z_width = scale(Sponge_width_cm)[,1],
    z_grp   = scale(Group_size)[,1],
    z_TL    = scale(TL_mm)[,1]
  )

d.alt<- d%>%
  filter(!is_male == "na") #208 individuals left

summary(d)
table(d$Rank, useNA="ifany")
table(d$sex, useNA="ifany")
```

```{r}
sex_str<- ggplot(d.alt, aes(x = TL_mm, fill = sex)) +
  geom_histogram(position = "dodge", bins = 15, color = "white") +
  scale_fill_manual(values = c(female = "#950952", male = "#104547")) +
  labs(x = "Total length (mm)",
       y = "Number of Individuals",
       fill = "Sex") +
  theme_classic()+
  theme(legend.position = "top")

sex_str

ggsave(file.path(output_dir, "Figure_4A.pdf"), plot = sex_str, width = 5, height = 4, units = "in")
```

How do social rank, sex, and habitat size together shape individual body size?

```{r}
#count individuals per rank x sex
sex_rank_df <- d.alt %>%
  mutate(
    Rank_n = as.numeric(as.character(Rank)),
    sex    = tolower(sex)
  ) %>%
  filter(Rank_n %in% 1:7, sex %in% c("male","female")) %>%
  count(Rank_n, sex, name = "n") %>%
  group_by(sex) %>%                           
  mutate(
    total_sex = sum(n),
    prop_sex  = n / total_sex
  ) %>%
  ungroup()

sex_rank_df
```

```{r}
#side-by-side bars per rank
sex_prop<-ggplot(sex_rank_df, aes(x = Rank_n, y = prop_sex, fill = sex)) +
  geom_col(position = position_dodge(width = 0.85), width = 0.8, color = "white") +
  scale_y_continuous(labels = percent_format(accuracy = 1),
                     limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = 1:7) +
  scale_fill_manual(values = c("female" = "#950952", "male" = "#104547"),
                    name = "Sex") +
  labs(
    x = "Social rank",
    y = "Proportion of individuals within sex"
  ) +
  theme_classic(base_size = 13)+
  theme(legend.position = "none")

sex_prop

ggsave(file.path(output_dir, "Figure_4B.pdf"), plot = sex_prop, width = 6, height = 4, units = "in")

```

### Sex, social rank, and body size

This linear mixed model tests whether males and females differ in body size at comparable ranks while controlling for sponge width and repeated observations within sponges (Supplementary Figure S10).

```{r}
d.alt <- d.alt %>%
  mutate(Rank_n = as.numeric(as.character(Rank)))

d.alt.f <- d.alt %>%
  filter(Rank_n <= 7)

m_add <- lmer(z_TL ~ Rank + sex + z_width + (1|Sponge_id), data = d.alt.f, REML = TRUE)
summary(m_add)

```

Interpreation:

Basic what is shown by the SEM
- There’s a steep size hierarchy: by Rank 3, individuals are ~3 TL units smaller than Rank 1 (controlling for width, sex).
- Males are larger than females across ranks; the shape of the hierarchy doesn’t differ by sex (interactions unsupported).
- Habitat size scales body size for everyone.

```{r}
#SD of raw width (cm): this defines what "1 SD" means
sd_width <- sd(d.alt.f$Sponge_width_cm, na.rm = TRUE)

#Pull the standardized slope from the model directly
b_z <- fixef(m_add)[["z_width"]]

#Odds ratios
OR_1SD <- exp(b_z)                 # per 1 SD increase in width
OR_1cm <- exp(b_z / sd_width)      # equivalent OR per 1 cm increase

sd_width
OR_1SD
OR_1cm
```

```{r}
#Model-based predictions for every Rank × sex combo
emm_all <- emmeans(
  m_add,
  ~ Rank | sex,
  at = list(z_width = 0),   
  type = "response"
) %>% as.data.frame()

#Which combos were actually observed?
cells <- d.alt.f %>% count(Rank, sex, name = "n")

plot_df <- emm_all %>%
  left_join(cells, by = c("Rank","sex")) %>%
  mutate(
    obs = ifelse(is.na(n) | n == 0, "unobserved", "observed"),
    rank_num = as.numeric(as.character(Rank))
  )

```

```{r}
#plotting model predictions
pd <- position_dodge(width = 0.35)

sz_rank<- ggplot(plot_df, aes(x = rank_num, y = emmean, color = sex)) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL),
                width = 0, position = pd, linewidth = 1) +
  geom_point(aes(shape = obs), position = pd, size = 4, stroke = 2, alpha = 0.7) +
  scale_shape_manual(values = c(observed = 16, unobserved = 1)) +  # solid vs hollow
  scale_color_manual(values = c(female = "#950952", male = "#104547")) +
  scale_x_continuous(breaks = sort(unique(plot_df$rank_num))) +
  labs(x = "Social Rank", y = "Predicted Total Length (mm)", color = "Sex") +
  theme_classic(base_size = 13)+
  theme(legend.position = "top")

sz_rank

ggsave(file.path(output_dir, "Figure_S10.pdf"), plot = sz_rank, width = 5, height = 4, units = "in")

```

### Reproductive status and egg production

Female body size and sponge width are evaluated as predictors of egg presence and egg number (Figures S11, 4B 4C)

```{r}
#keeping only females
fem<- social.all %>%
  filter(is_male == 0)
fem$Rank<- as.factor(fem$Rank)
```

```{r}
#organizing dataset
fem<- fem %>%
  mutate(
    has_eggs = factor(eggs, levels = c(0,1),
                      labels = c("no_eggs", "eggs")))
fem_egg<- fem %>%
  filter(has_eggs == "eggs")

sum<- fem_egg%>%
  summarize(
    eggs_avg = mean(egg_count)) #obtaining counts for eggs

fem_per<- fem %>%
  count(has_eggs) %>%
  mutate(percent = 100 * n / sum(n))

```

#### Gravid females across social ranks

Female total length is plotted by social rank and colored by egg status. This descriptive plot reproduces Supplementary Figure S11 and shows that gravid females are not restricted to rank 1.

```{r}
egg_status<- ggplot(fem, aes(x = Rank, y = TL_mm, color = has_eggs)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = c("#9A6D38","#F4D35E")) +
  labs(x = "Rank",
       y = "Total length (mm)",
       color = "Eggs") +
  theme_classic()

egg_status

ggsave(file.path(output_dir, "Figure_S11.pdf"), plot = egg_status, width = 5, height = 4, units = "in")

```

Are females carrying eggs significantly larger?

- Does female body size predict reproduction?

- Does habitat size (sponge size) further predict reproduction, beyond body size?

It lets you conclude whether habitat capacity influences whether females reproduce, not just how much they produce.

Bigger sponges could:

- allow more food / better flow → higher condition → more eggs, or

- simply allow more fish but not change per-female reproduction.

```{r}
#modeling the presence of eggs as a function of body size and sponge size
m_eggs_presence <- glmmTMB(
  has_eggs ~ scale(TL_mm) + scale(Sponge_width_cm) + (1 | Sponge_id),
  family = binomial,
  data = fem
)

summary(m_eggs_presence)

#sd(fem$TL_mm, na.rm = TRUE)
```

Only female body size has a direct effect on the presence of eggs.

```{r}
#Visualizing eggs status - rainclound plot 
ggplot(fem, aes(x = has_eggs, y = TL_mm, fill = has_eggs)) +
  stat_halfeye(
    adjust = .6,
    width = .5,
    .width = 0,
    justification = -.2,
    alpha = .6,
    color = NA
  ) +
  geom_boxplot(
    width = .2,
    outlier.shape = NA,
    alpha = .7
  ) +
  geom_jitter(
    aes(color = has_eggs),
    width = .08,
    size = 1.4,
    alpha = .8
  ) +
  
  scale_fill_manual(values = c("grey40","#F4D35E")) +
  scale_color_manual(values = c("grey40","#F4D35E")) +
  
  labs(y = "Total length (mm)") +
  theme_classic(base_size = 12) +
  theme(legend.position = "none", axis.title.x = element_blank())

ggsave(file.path(output_dir, "Figure_4C.pdf"), width = 5, height = 4, units = "in")
```

Females carrying eggs were significantly larger and sponge size has no direct effect.

- But are the bigger females having more eggs (egg count)?

```{r}
#modeling egg count as a function of body size and sponge size
#count data - overdisperse - negative binomial
m_egg_count <- glmmTMB(
  egg_count ~ scale(TL_mm) + scale(Sponge_width_cm) + (1 | Sponge_id),
  family = nbinom2,
  data   = fem_egg)

summary(m_egg_count)

#sd(fem$TL_mm[fem$has_eggs == "eggs"], na.rm = TRUE)
```

```{r}
pred_egg <- ggpredict(m_egg_count, terms = "TL_mm [all]")
```

```{r}
egg_count<- ggplot() +
  geom_ribbon(
    data = pred_egg,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    fill = "#F4D35E",
    alpha = 0.2
  ) +
  geom_line(
    data = pred_egg,
    aes(x = x, y = predicted),
    linewidth = 1
  ) +
  geom_point(
    data = fem_egg,
    aes(x = TL_mm, y = egg_count),
    colour = "#F4D35E",
    size = 4,
    alpha = 0.85
  ) +
  labs(
    x = "Total length (mm)",
    y = "Egg count"
  ) +
  theme_classic(base_size = 13)

egg_count

ggsave(file.path(output_dir, "Figure_4D.pdf"), plot = egg_count, width = 5, height = 4, units = "in")
```

### Age structure and body size

The otolith-aged subset is used to summarize age distributions and test whether age, sex, dominant status, group size, and sponge width explain body size.

```{r}
oto<- social.all %>%
  filter(!is.na(age)) #31 individuals

summary(oto$age)
head(oto)
```

```{r}
#organizing otolith data 
oto <- transform(oto,
  Age_c = scale(age, center = TRUE, scale = FALSE),
  Group_size_c = scale(Group_size, center = TRUE, scale = FALSE),
  Sponge_c = scale(Sponge_width_cm, center = TRUE, scale = FALSE)
)

oto<- oto %>%
  mutate(Rank_num = as.numeric(as.character(Rank)))

oto<- oto %>%
    mutate(
    Rank1 = ifelse(Rank_num == 1, "Rank1", "NotRank1"),
    Rank1 = factor(Rank1, levels = c("NotRank1", "Rank1")))
  
oto<- oto %>%
  mutate(
    sex = case_when(
      str_to_lower(is_male) %in% c("male","m","1","true","t") ~ "male",
      str_to_lower(is_male) %in% c("female","f","0","false","fa") ~ "female",
      TRUE ~ NA_character_
    ) %>% factor(levels = c("female","male")))


```

```{r}
#Obtaining basic metrics

age_summary <- oto %>%
  summarise(
    n_age      = n(),
    mean_age   = mean(age, na.rm = TRUE),
    sd_age     = sd(age, na.rm = TRUE),
    min_age    = min(age, na.rm = TRUE),
    max_age    = max(age, na.rm = TRUE),
    mean_TL    = mean(TL_mm, na.rm = TRUE),
    sd_TL      = sd(TL_mm, na.rm = TRUE),
    min_TL     = min(TL_mm, na.rm = TRUE),
    max_TL     = max(TL_mm, na.rm = TRUE)
  )
age_summary
```

```{r}
age_by_sex <- oto %>%
  group_by(sex) %>%
  summarise(
    n          = n(),
    mean_age   = mean(age, na.rm = TRUE),
    sd_age     = sd(age, na.rm = TRUE),
    mean_TL    = mean(TL_mm, na.rm = TRUE),
    sd_TL      = sd(TL_mm, na.rm = TRUE),
    .groups = "drop"
  )
age_by_sex

```

```{r}
age_by_rank1 <- oto %>%
  group_by(Rank) %>%
  summarise(
    n          = n(),
    mean_age   = mean(age, na.rm = TRUE),
    sd_age     = sd(age, na.rm = TRUE),
    mean_TL    = mean(TL_mm, na.rm = TRUE),
    sd_TL      = sd(TL_mm, na.rm = TRUE),
    .groups = "drop"
  )
age_by_rank1

```

```{r}
#Obtaing age distribution and frequency
d_oto <- oto %>%
  dplyr::rename(Age_days = age)

#Pivot to long format
df_long <- d_oto %>%
  pivot_longer(
    cols = c(Age_days, TL_mm),
    names_to = "Measure",
    values_to = "Value"
  ) %>%
  mutate(
    Measure = recode(Measure,
                     Age_days = "Age (days)",
                     tL_mm = "Total length (mm)"),
    Units = if_else(Measure == "Age (days)", "days", "mm")
  )

#Calculate means for dashed line + label
means <- df_long %>%
  group_by(Measure, Units) %>%
  summarise(m = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  mutate(lbl = sprintf("Mean: %.1f %s", m, Units))

#Visualizing age distribution
ggplot(df_long, aes(x = Value)) +
  geom_histogram(fill = "grey", color = "white", bins = 15) +
  facet_wrap(~Measure, scales = "free_x", nrow = 1, strip.position = "bottom") +
  geom_vline(data = means, aes(xintercept = m),
             linetype = "dashed", linewidth = 0.7) +
  geom_text(
    data = means,
    aes(x = m, y = Inf, label = lbl),
    vjust = 1.2, size = 3.5, fontface = "bold"
  ) +
  labs(x = NULL, y = "Frequency") +
  theme_classic(base_size = 12) +
  theme(
    #strip.placement = "outside",       # move strips outside plot panels
    #strip.position = "bottom",         # put them below the x-axis
    strip.text = element_text(face = "bold"),
    strip.background = element_blank()
  )
ggsave(file.path(output_dir, "Figure_5A.pdf"), width = 5, height = 4, units = "in")
```

#### Do dominant individuals exhibit distinct age or size characteristics? (Figure 5B)

```{r}
# The corresponding mixed models had a zero Sponge_id random-effect variance
# (singular fits). Ordinary linear models therefore retain the supported
# fixed-effect hypotheses without interpreting an unsupported random effect.
fm0 <- lm(TL_mm ~ Age_c + Group_size + Sponge_c, data = oto)
fm1 <- lm(TL_mm ~ Age_c + sex + Rank1 + Group_size + Sponge_c, data = oto)
fm2 <- lm(TL_mm ~ Age_c * Rank1 + sex + Group_size + Sponge_c, data = oto)
fm3 <- lm(TL_mm ~ Age_c * sex + Rank1 + Group_size + Sponge_c, data = oto)
fm4 <- lm(TL_mm ~ Age_c * sex + Age_c * Rank1 + Group_size + Sponge_c, data = oto)
fm5 <- lm(TL_mm ~ Age_c * sex * Rank1 + Group_size + Sponge_c, data = oto)

aictab(list(fm0,fm1,fm2,fm3,fm4,fm5),
       modnames = c("Age","Age+sex+rank","Age*rank","Age*sex","Age*sex+Age*rank","3way"))
```

```{r}
summary(fm1) #best-supported non-singular model
```

```{r}
#Organize predicted data for plotting
pred_age <- ggpredict(fm1, terms = "Age_c [all]")

age_mean <- mean(oto$age, na.rm = TRUE) 
pred_age$age <- pred_age$x + age_mean

oto_age<- ggplot(oto, aes(x = age, y = TL_mm)) +
  geom_ribbon(
    data = pred_age,
    aes(x = age, ymin = conf.low, ymax = conf.high),
    inherit.aes = FALSE,
    alpha = 0.2
  ) +
  
  geom_line(
    data = pred_age,
    aes(x = age, y = predicted),
    inherit.aes = FALSE,
    linewidth = 1
  ) +
  
  geom_point(
    aes(color = sex, shape = Rank1),
    size = 4,
    alpha = 0.7
  ) +
  
  scale_color_manual(values = c(female = "#950952", male = "#104547")) +
  scale_shape_manual(values = c(NotRank1 = 16, Rank1 = 18)) +
  
  labs(
    x = "Age (days)",
    y = "Total length (mm)",
    color = "Sex",
    shape = "Rank"
  ) +
  theme_classic(base_size = 13) +
  theme(legend.position = "top")

oto_age

ggsave(file.path(output_dir, "Figure_5B.pdf"), width = 5, height = 4, units = "in")

```

#### Does *Risor ruber* age vary across sampled sponges?

Age, sex, and dominant status are displayed for each of the four sponges represented in the otolith subset. The expected output is Supplementary Figure S12; unidentified sex is retained as `NA` in the figure.

```{r}
summ_sponge<-oto %>%
  group_by(Sponge_id) %>%
  summarise(
    n_total   = n(),
    n_female  = sum(sex == "female", na.rm = TRUE),
    n_male    = sum(sex == "male", na.rm = TRUE),
    age_min   = min(age, na.rm = TRUE),
    age_max   = max(age, na.rm = TRUE),
    age_mean  = mean(age, na.rm = TRUE),
    .groups   = "drop"
  )

#write.csv(summ_sponge, "mean_age_sponge_supp.csv")
```

```{r}
oto2 <- oto %>%
  mutate(
    Sponge_id = fct_reorder(Sponge_id, age, .fun = mean, na.rm = TRUE))

#visualizing age variation across sampled sponges
ggplot(oto2, aes(x = age, y = Sponge_id)) +
   geom_point(aes( colour = sex, shape = Rank1),
    size = 4,
    alpha = 0.7
  ) +
    scale_color_manual(values = c(female = "#950952", male = "#104547")) +
  scale_shape_manual(values = c(NotRank1 = 16, Rank1 = 18)) +
  scale_size_continuous(name = "Total Length (mm)") +   # legend
  labs(x = "Age (days)", y = "Sponge (group)", colour = "Sex") +
  theme_classic()

ggsave(file.path(output_dir, "Figure_S12.pdf"), width = 5, height = 4, units = "in")
```

## Session information

The following output records the R, operating-system, and package versions used to render this report.

```{r session-information}
sessionInfo()
```
